Reproducible and programmatic generation of neuroimaging visualisations

### 2022-04-05 manuscript.Rmd
fig update

The following four sections provide brief examples packages and
functions available in R for visualising voxel, vertex, ROI and
edge-level data. A more didactic introduction to these tools is provided
in a [collection of beginner friendly
R-Notebooks](https://github.com/sidchop/neuRo-vis-BrainHack-2021).

\newpage
\blandscape

### Figure 1 [Fix image size ratios]

@@

#loop this?
fig_name <- c('fig1a', 'fig1b', 'fig1c', 'fig1d', 'fig1e', 'fig1f', 'fig1g1', 'fig1g2', 'fig1g3','fig1h', 'fig1i')
figs <- list()

@@

patchwork <- figs[[1]]/figs[[2]]|(figs[[3]]|figs[[6]])/figs[[4]]|(figs[[5]]|(figs[[8]]/figs[[9]]))/figs[[7]]|figs[[10]]/figs[[11]]

@@

The `oro.nifti` package allows for the reading, writing, manipulation
and visualisation of voxel-level imaging data of either nifti, analyze
or afni formats. In particular, the `image`, `slice` and `orthographic`
functions allow users to display the desired number of slices in the
desired orientations, while also providing precise control over overlaid
images, colour scales, legend placement and other aesthetic settings
(Fig 1A). Smoothing and re-sampling of images is sometimes required for
visualisation purposes, which can be achieved via the `ANTsR`
/`extrantsr` or `fslr` packages.

@@

The recently developed `fsbrain` package allows for the visualisation of
vertex-level and atlas-level data obtained using Freesurfer. It contains
many flexible functions for visualising individual-level and group-level
measures such as cortical thickness, volume or surface area (Fig 1B).
The package also comes with extensive guides and documentations to
assist users (cite).

CIFTI or 'grey-ordinate' data is becoming a popular format for
structural and functional imaging data, as it combines vertex-level data
of the cortex with voxel-level data fo the cerebellum and sub-cortex. In
R, CIFTI data can now be read, visualised and manipulated using the
well-documented `ciftiTools` package (Fig 1C; cite).

@@

Statistical parameters can be mapped onto discreet cortical and
sub-cortical regions using the `ggseg` and `ggseg3d` package. These
packages flexibly generates aesthetic renderings of cortical,
sub-cortical and white matter ROIs in 2D and 3D (Fig 1D). Usually,
ROI-level interpretations are dependent on a standardised atlas or
parcellation schemes the brain, accordingly this package and its add-ons
(`ggExtra`) provides a large array of commonly used atlases and some
functionality for users to contribute other atlases. Additionally, being
part of the 'Grammar of Graphics' frame work allows for integration with
packages such as `gganimate`, enabling users to animate and visualise
dynamic changes in spatial and temporal statistics across ROIs. The
`ciftiTools` package also allows for surface and voxel-level
visualisation of ROIs for the cortex and sub-cortex respectively (Fig
1E). Flexibly allowing CIFTI format atlases and ROIs to be displayed.

@@

# ADD COMMENTS TO EXPLAIN EACH LINE

library(neurobase)
library(ggplotify)

@@

png("data/fig1a.png") 
ortho2(x = template, 
       y = effect, 
       crosshairs = F, 
       bg = "white", 
       NA.x = T, 
       col.y= hotmetal(), 
       xyz = c(70,50,80),
       useRaster = T,
       ycolorbar = TRUE, 
       mfrow = c(1,3)) +
  colorbar(breaks,col = hotmetal(n=length(breaks)-1), 
           text.col = "black", labels = TRUE, maxleft = 0.95) 
dev.off()

@@

```{r Fig1c , eval=FALSE}
#vertex 1
#Figure 1B
library(rgl)
library(ggplot2)
# Load the package and point to the Connectome Workbench
suppressPackageStartupMessages(library(ciftiTools))

@@

cifti_fname <- ciftiTools::ciftiTools.files()$cifti["dtseries"]
surfL_fname <- ciftiTools.files()$surf["left"]

cii <- read_cifti(
  cifti_fname, brainstructures="left", 
  surfL_fname=surfL_fname)

@@

#weird angle

@@

#download_optional_data()
subjects_dir = get_optional_data_filepath("subjects_dir")

subject_id = 'subject1'

colourmap = colorRampPalette(viridis::viridis(n = 100))

vis.subject.morph.native(subjects_dir = subjects_dir, 
                         subject_id = 'subject1', 
                         measure = 'thickness', 
                         hemi='both', 
                         views = 't9',
                         draw_colorbar = "horizontal",
                         cortex_only = T, surface = "white", 
                         makecmap_options = list('colFn'=colourmap))
rgl.snapshot("data/fig1d.png")
rgl.close()

@@

plot2 <- view_xifti_surface(xii_metric, 
                            colors = viridis::viridis(n = 400), 
                            borders = "black", 
                            hemisphere = "left", )

@@

subcortex <- ggseg(atlas = aseg,
                   .data = sub_base_atlas,
                   hemisphere = "right",
                   mapping=aes(fill=Effect), 
                   position="stacked",
                   colour="black",
                   size=.2,
                   show.legend = T, 
                   plot.background = "white") + 
  scale_fill_viridis_b()

@@

set.seed(1993)
volume <- sample(1:20, 14) #randomly generated data for 14 structures
make_ggseg3d(volume, 
             colour.pal = c("white", "firebrick"), 
             output.png = T, file.name = 
               "data/fig1g3.png")

>>>>

The following four sections provide brief examples packages and
functions available in R for visualising voxel, vertex, ROI and
edge-level data. A didactic introduction to these tools is provided in a
accompaning [collection of beginner friendly
R-Notebooks](ttps://github.com/sidchop/neuRo-vis-BrainHack-2021).

```{=tex}
\newpage
\blandscape
```
### Figure 1 [Fix image size ratios]

@@

fig_name <- c('fig1a', 'fig1b', 'fig1c', 'fig1d', 'fig1e', 'fig1f', 'fig1g1', 'fig1g2', 'fig1g3','fig1h', 'fig1i')
figs <- list()

@@

patchwork <- figs[[1]]/(figs[[2]]|figs[[3]])|(figs[[6]]/(figs[[8]]|figs[[9]]))/figs[[4]]|figs[[10]]/figs[[11]]

@@

The `oro.nifti` package allows for the reading, writing, manipulation
and visualisation of voxel-level imaging data of either nifti, analyze
or afni formats. In particular, the `image`, `slice` and `orthographic`
functions allow users to display the desired number of slices in the
desired orientations, while also providing precise control over overlaid
images, colour scales, legend placement and other aesthetic settings
(Fig 1A-B). Smoothing and re-sampling of images is sometimes required
for visualisation purposes, which can be achieved via the `ANTsR`
/`extrantsr` or `fslr` packages.

@@

The recently developed `fsbrain` package allows for the visualisation of
vertex-level and atlas-level data which is derieved using FreeSurfer
formats. It contains many flexible functions for visualising
individual-level and group-level measures such as cortical thickness,
volume or surface area (Fig 1E). The package also comes with extensive
guides and documentations to assist users (cite).

CIFTI or 'grey-ordinate' data is becoming a popular format for
structural and functional imaging data, as it combines vertex-level data
of the cortex with voxel-level data for the cerebellum and sub-cortex.
In R, CIFTI data can now be read, visualised and manipulated using the
well-documented `ciftiTools` package (Fig 1C-D,F; cite).

@@

Statistical parameters can be mapped onto discreet cortical and
sub-cortical regions using the `ggseg` and `ggseg3d` package. These
packages flexibly generates aesthetic renderings of cortical,
sub-cortical and white matter ROIs in 2D and 3D (Fig 1D). Usually,
ROI-level interpretations are dependent on a standardised atlas or
parcellation schemes the brain, accordingly this package and its add-ons
(`ggExtra`) provides a large array of commonly used atlases and some
functionality for users to contribute other atlases. Additionally, being
part of the 'Grammar of Graphics' frame work allows for integration with
packages such as `gganimate`, enabling users to animate and visualise
dynamic changes in spatial and temporal statistics across ROIs.

The `ciftiTools` package also allows for surface and voxel-level
visualisation of ROIs for the cortex and sub-cortex respectively (Fig
1E). Flexibly allowing CIFTI format atlases and ROIs to be displayed.

@@

# ADD COMMENTS TO EXPLAIN EACH LINE
library(neurobase)
library(ggplotify)

@@

png("data/fig1a.png") 
ortho2(x = template, 
       y = effect, 
       crosshairs = F, 
       bg = "white", 
       NA.x = T, 
       col.y= viridis::viridis(n = 500), 
       xyz = c(70,50,80),
       useRaster = T,
       ycolorbar = TRUE, 
       mfrow = c(1,3)) +
  colorbar(breaks,col =viridis::viridis(n=length(breaks)-1), 
           text.col = "black", labels = TRUE, maxleft = 0.95) 
dev.off()

@@

```{r Fig1c , eval=FALSE}
#vertex 1
#Figure 1B
library(rgl)
library(ggplot2)
# Load the package and point to the Connectome Workbench
library(ciftiTools)

@@

cifti_fname <- ciftiTools::ciftiTools.files()$cifti["dtseries"]
surfL_fname <- ciftiTools.files()$surf["left"]
surfR_fname <- ciftiTools.files()$surf["right"]

cii <- read_cifti(
  cifti_fname, brainstructures="all", 
  surfL_fname=surfL_fname,
  surfR_fname=surfR_fname)

@@

@@

#download_optional_data()
#download_fsaverage(accept_freesurfer_license = TRUE)

subjects_dir <- get_optional_data_filepath("subjects_dir")

subject_id <- 'subject1'

colourmap <- colorRampPalette(viridis::viridis(n = 10000))
fsbrain.set.default.figsize(700, 700);

cm = vis.subject.morph.standard(subjects_dir, 
                                subject_id, 'sulc', 
                                fwhm='10', 
                                cortex_only = T, 
                                views=NULL, 
                                makecmap_options = list('colFn'=colourmap))
img = export(cm, colorbar_legend='Effect', output_img = "data/fig1d.png")

rgl.snapshot("data/fig1d.png")
rgl.close()

@@

plot2 <- view_xifti_surface(xii_metric, 
                            colors = viridis::viridis(n = 400), 
                            borders = "black", 
                            hemisphere = "both",
                            view = 'lateral')

@@

subcortex <- ggseg(atlas = aseg,
                   .data = sub_base_atlas,
                   mapping=aes(fill=Effect), 
                   position = "dispersed",
                   colour="black",
                   hemi =c('left', 'right'),
                   size=.2,
                   show.legend = T, 
                   plot.background = "white") + 
  scale_fill_viridis_b()

@@

set.seed(1993)
volume <- sample(1:20, 14) #randomly generated data for 14 structures
make_ggseg3d(volume, 
             colour.pal = viridis::viridis(n=length(volume)), 
             output.png = T, file.name = "data/fig1g3.png")
